---
title: "powergrid"
output:
 rmarkdown::html_vignette:
  toc: true
  toc_depth: 3
vignette: >
  %\VignetteIndexEntry{powergrid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=100
)
```

```{r setup}
library(powergrid)
```

# Introduction to the typical use of powergrid
The powergrid package offers a number of functions that facilitate
exploring the statistical power of a study. In a typical use case
(which I also show in this vignette) you want to explore the relation
between statistical power, sample size, the effect size of interest
and the assumed variability of the data. The interrelation of these
parameters is defined by the assumed distributional form of your data,
and the statistical test that you aim to perform.

This package does not dictate any of the above ingredients for you. It
just allows you to evaluate a function at a grid of parameters and
visualize the interrelation in plots that are fine-tuned for analyses
of power and sample size. You may, however use it for very different
purposes. An example close to a power analysis would be an analysis of
the precision you may expect of a study design. As a more remote
application, you can think of a coverage study for different
confidence intervals. Or an exploration for planning recruitment under
various scenarios.

## Functions covered
T his vignette covers the following functions of `powergrid`

- `PowerGrid` for evaluating the power across a grid, with and without simualtion
- `Example` for inspecting the relation between power and sample size
  in a specific scenario
- `Powerplot` Graphical exploration of power under different scenarios
- `GridPlot` Graphical exploration of power under even more scenarios

# Typical use case
I describe two basic use cases here. Both concern studying the relation between
power, sample size, and two further parameters. In the first situation, a
function for calculating the power is available. In the second situation, the
power needs to be found through (re-)sampling.

## Power for a t-test on a normal variable
Assume we aim to collect data in a two-armed RCT and plan to perform a simple
t-test. In this situation, the situation concerning power can be summarized in
the following ingredients:

1. total sample size
2. effect size of interest
3. expected standard deviation in the study arms
4. the objective (planned statistical test)

I use the function `PowerGrid` to evaluate the situation sketched
above. This is done as follows

```{r, basic PowerGrid}
sse_pars = list( # a simple list
  n = seq(from = 10, to = 60, by = 5), # sample size
  delta = seq(from = 0.6, to = 1.6, by = 0.1), # effect size
  sd = seq(.5, 1.5, .2) # variability
)

PowFun <- function(n, delta, sd){
  ptt = power.t.test(n = n/2,
                     delta = delta,
                     sd = sd,
                     sig.level = 0.05)
  return(ptt$power)
}

power = PowerGrid(pars = sse_pars, fun = PowFun)
summary(power)

```

In the code above, the first 3 ingredients are prepared as a little list of
parameters `pars`. The third ingredient, the objective, is prepared as a
function `PowFun`. Note these are simple base R objects. Note further,
that the names of the elements in the list `pars` match the names of the
function arguments in PowFun. 

Once that is set, in the third part of the chunk above, PowerGrid is called with
the sse\_pars and the PowFun for arguments pars and fun, respectively. The result
is an array with class `power\_array`. You can use it as any array, but
it is built to be used in some neat functions of PowerGrid.

## Finding n for a certain set of assumptions
First, you may want to calculate the minimal n required to achieve, for example,
a power of 90\%, for detecting a effect size of 0.8, assuming the standard
deviation in your sample will be 0.9. To find this n, use the function
`Example`:

```{r, basic Example}
Example(power, # the object (power_array) I created above
        example = list(delta = .8, sd = .9), # at which settings do I want to
                                             # calculate n?
        target = .9) # what is my target power?
```

The printed result of Example should speak for itself. Note, however, that the
wording "target" in both the function argument and the printed result may be a
bit vague. 'Why not "power"?', you may ask. The reason is, that there is nothing
that keeps you from optimizing other things with the functions in
`powergrid`. Indeed, instead of finding a target power, you may be looking for a
target precision.

If you save the output of Example, this is a list containing the relevant
information in a reusable form:

```{r basic Example list}
sse_example = Example(power,  example = list(delta = .8, sd = .9), target = .9)
str(sse_example) # a list
```

## Exploring the uncertainty about power
The example that was calcaulted above does not reflect the uncertainty
about my study. I set sd to .9, but I do not really know what SD to
expect. And I may want to explore how my power depends on the effect
size I want to detect. To explore the uncertainty about power in a
study design study, `powergrid` offers two plotting functions:
`PowerPlot` and `GridPlot`.

### PowerPlot

First, I can plot the required n for different effect sizes and a
chosen SD to achieve a certain power.

Below, I show how to do this when I aim to focus on the assumption that the
standard deviation equals .9. In the example below, I want to point at the
effect size that I hope to be able to show, which is 1. The figure shows
"isolines" connecting points where power is .8, .9, and .95.

```{r, basic PowerPlot, fig.width=6, fig.height=6, fig.align='center'}
PowerPlot(
  power, # the power_array containing power estimates
  slicer = list(sd = .9), # "cut out" the relevant slice from power.
  example = list(delta = 1) # draw an arrow pointing at the situation at delta = 1
  )
```

Some things to note: 

* You need to "slice out" one plain from your power_array. In this case, this
  the slice where `sd = .9`. The slice has the form of the figure: delta
  by n.
* There is nothing that keeps you from slicing out a plain where `delta = .8`
  and show how the relation between power and n depends on the standard
  deviation. Or slicing out a plain where `n = 50` (maybe that's considered
  feasible for your study project), and see how power behaves as a function of
  delta and sd.
* You can add additional examples, either by increasing the the length of the
  vector in the argument to example, (e.g., `example = list(delta = c(1, 1.2)))`
  or by using the higher level plotting function `AddExample`. The latter allows
  you more flexibility, like setting different colors or line types.
* There are many options in PowerPlot and AddExample that you may want to learn
  about in the help files.

### GridPlot
The figure created by `PowerPlot` can only show the interplay of two variables
and power. GridPlot often offers a more insightful picture, in particular when,
as in this example, we have more than 2 dimensions in our `pars` argument.

The code below shows how to plot the interplay between n, delta and sd when the
goals is the achieve 90% power.

```{r, GridPlot, fig.width=6, fig.height=6, fig.align='center'}
GridPlot(power,
         target = .9, # you need to choose one target level of power
         example = list(delta = 1, sd = .9)) # defined by two parameters now.
```

Note that there are many options in this plot. See the help file of `GridPlot`
for more info.

## Power evaluation using simulation and resampling
Assume we have about the same situation as above, but we do not have a
simple solution to calculate the power: we only have a limited pilot data set
that looks as follows:
```{r, pilot data}
pilot_scores = c(2.1, 4.3, 2.3, 5.2, 1.9, 8.3, 7, 2.6, 2.4, 3.2, 2.1, 2.8, 3.4)
```

Since we do not really understand the distribution (it looks pretty
right-skewed), we plan to perform a Mann-Whitney U-test. We do not want to
simply simulate, but draw from our pilot sample to mimic the variability and
distributional form. We do have a clear idea about effect size (somewhere in the
range of .5 and 2). The following code my be our approach to the exploration of
power:

```{r, resample MannU, fig.width=6, fig.height=6, fig.align='center', warning = FALSE}
sse_pars = list(
  n = seq(10, 100, 20),
  delta = seq(.5, 2, .2)) # only effect size
PowFun = function(n, delta, pilot_data){
  arm_1 = sample(pilot_data, n, replace = TRUE)
  arm_2 = sample(pilot_data, n, replace = TRUE) + delta
  significant = wilcox.test(arm_1, arm_2)$p.value < .05
  return(significant) # each call of this function gives significant either TRUE
                      # or FALSE
  }
power = PowerGrid(pars = sse_pars,
                  fun = PowFun,
                  more_args = list(pilot_data = pilot_scores), # pass the pilot
                                                               # data on to the
                                                               # fun argument
                  n_iter = 99) # we need to iterate over simulated experiemtns
                               # to get a power. I would take a higher value
                               # than 99; this is to keep the example quick.
summary(power)
PowerPlot(power)
```

A couple of notes:

* The power in the example above is calculated by simulating TRUE's and FALSE's
  for significance. These were then automatically summarized by `mean` to yield
  the power. You may, however, want to keep the simulated outcomes of your
  function. To do so, set `summarize = FALSE`. If you choose to keep the
  individual iterations, be aware that plotting functions and `Example`
  automatically summarize these taking the mean. You can, however, choose a
  different `summary_function`.
* There are a range of graphical options, including the labeling of axes and
  lines.
* As in PowerPlot, any dimension of the `power_array` may be represented by the
  x-axis, y-axis, and lines.



